\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-19 coqimpl.tex
Save language design

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy can allow fields to have type dependent on former fields thus fulfilling \ref{langdesign-req1}. However, there are some caveats, especially when the form fields are actually ``proof-relevant''. For example, we may define a field $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to proving its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. A more concrete example is in \cref{fig:plugin-example1}, consider we have "Field b : self_A.a = 1 := .." instead, then our compilation and Coq's type check will both fail: in the context of "b" (module type "a_6") we only know "a" is of type nat but nothing more (and this is necessary so that "a" can be overridden in the future). To fix that, we need more information on "a" than its type---we need to expose its concrete definition "a = 1" to appear in the context of "b".

@@

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is not expressive enough to really fulfill \ref{langdesign-req2}. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. Thus, apart from the (total) recursor, we need to have another formulation of the recursor i.e., we need to have a better characterization of ``extensible inductive type'' exported in the module type as interface.

@@

\textbf{Hack: ``Meta-data''.}

>>>>

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy can allow fields to have type dependent on former fields thus fulfilling \ref{langdesign-req1}. However, there are some caveats, especially when the former fields are actually ``proof-relevant''. For example, we may define $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to prove its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. A more concrete example is in \cref{fig:plugin-example1}, consider the case we have "Field b : self_A.a = 1 := .." instead, then our compilation and Coq's type check will both fail: in the context of "b" (module type "a_6") we only know "a" is of type nat but nothing more (and this is necessary so that "a" can be overridden in the future). To fix that, we need more information on "a" than its type---we need to expose its concrete definition "a = 1" to appear in the context of "b".

@@

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is not expressive enough to really fulfill \ref{langdesign-req2}. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor of (vanilla) inductive type, we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. Thus we need to alter the formulation of the recursor i.e., we need to have a better characterization of ``extensible inductive type'' exported in the module type as interface.

@@

\textbf{Hack: ``Meta-data''.} Due to our implementation of the family, the family data is invisible for Coq internal and thus cause some inconvenience.  For example, currently it is not possible to define a vanilla inductive type dependent on our ``extensible'' inductive type due to the invisiblity, unless we compile the whole family. Similarly, customized tactic expression requested by \ref{langdesign-req5} cannot use lemmas or theorems in the fields of the current defining family as well.

To remedy this, we facilitate a "MetaData" command. This command will allow to bundle any original Coq primitive (in the current family context) into the module type, and thus visible to the following defining field. Doing so, we can define customized tactic as in \ref{langdesign-req5} that uses lemmas proved in the family.

### 2022-06-19 lang-design.tex
Save language design

Let's start with the basic example---STLC and its extension---to consider what kind of features are required and how much of them can be supported by family polymorphism.
% add the example code here

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to reason about fields in a family}, just like in a module, where a field can reason about its former fields. More generally, we want it possible for a later field to have type dependent on the former fields.

% we want extensible inductive type
\textlabel{Req (2)}{langdesign-req2}:\textbf{ we want extensible inductive type}. The extension of inductive type is not supported by \citet{zm2017} thus we need further consideration about it on both implementation and metatheory. Additionally, \textlabel{Req (3)}{langdesign-req3}:\textbf{we also want extensible pattern matching and induction reasoning}. There are actually two kinds of ``pattern matching'', one for data and the other for induction reasoning, i.e. one use the eliminator to \mintinline{Coq}{Set} or \mintinline{Coq}{Type}, and the other uses the eliminator to \mintinline{Coq}{Prop}. Luckily, in this setting, extensible pattern matching can be easily expressed with family extension -- we just aggregate all case handlers of pattern matching into a family (as a bunch of functions), and then one family can encode one pattern matching, and family inheritance can express adding case handlers. Then we just need to introduce a primitive that will ``wrap'' that family into a recursive function. Induction reasoning can be handled in an almost identical way.

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction reasoning, because the later one is \textit{proof-irrelevant}. This difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:\textbf{we want to reason aboue extensible pattern matching}. Our substitution lemma is one example,

% we want tactic programming and certain level of automation
And finally, \textlabel{Req (5)}{langdesign-req5}:\textbf{we want tactic programming}.

>>>>

Let's start with the basic example---STLC and its extension in \cref{fig:STLC-example}---to consider what kind of features are required and how much of them can be supported by family polymorphism.
% add the example code here

\begin{figure}
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family STLC.
Inductive ty : Set :=
  | unit : ty
  | arrow : ty -> ty -> ty.
Inducitive term : Set := 
  | var : id -> term 
  | lam : id -> term -> term ...
Fixpoint subst 
  : id -> term -> term := ...
Inductive has_type 
  : context -> term -> type := ...
Proposition subst_lemma :
  has_type (Γ; x ↦ A) t T ->
  has_type Γ u A ->
  has_type Γ (subst x u t) T.
Inducitve step : term -> term -> Prop 
  := ...
(* ... and more, end with type safety *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.47\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family STLC_bool extends STLC.
Extend Inductive ty : Set :=
  | bool : ty.

Extend Inducitive term : Set := 
  | tt : term | ff : term 
  | tif : term -> term -> term -> term
Extend Fixpoint subst (* Need to handle new term *)
  : id -> term -> term := ...
Extend Inductive has_type (* .. and new ty too *)
  : context -> term -> type := ...
Extend Proposition subst_lemma :
  ... (* Need to prove extra cases *)

Extend Inducitve step : term -> term -> Prop 
  := ... (* Need to expand this binary relation *)
(* ... and more extension *)
EndFamily.
\end{minted}
  \end{minipage}
  \caption{Example STLC and its extension}\label{fig:STLC-example}
\end{figure}

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to reason about fields in a family}, just like in a module, where a field can reason about its former fields. More generally, we want it possible for a later field to have type dependent on the former fields. In the example of \cref{fig:STLC-example}, "subst", "has_type", "subst_lemma" all require this feature.

% we want extensible inductive type
\textlabel{Req (2)}{langdesign-req2}:\textbf{ we need extensible inductive type}, to extend "term" and "ty" in \cref{fig:STLC-example}. The extension of inductive type is not supported by \citet{zm2017} thus we need further consideration about it on both implementation and metatheory. Additionally, \textlabel{Req (3)}{langdesign-req3}:\textbf{we also want extensible pattern matching and induction reasoning}, to extend "subst" in \cref{fig:STLC-example}. There are actually two kinds of ``pattern matching'', one for data and the other for induction reasoning, i.e. one use the eliminator to \mintinline{Coq}{Set} or \mintinline{Coq}{Type}, and the other uses the eliminator to \mintinline{Coq}{Prop}. Luckily, in this setting, extensible pattern matching can be easily expressed with family extension -- we just aggregate all case handlers of pattern matching into one family (as a bunch of functions), and then one family can encode one pattern matching, and family inheritance can express adding case handlers. Then we just need to introduce a primitive that will ``wrap'' that family into a recursive function. Induction reasoning can be handled in an almost identical way.

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction reasoning, because the former one is \textit{proof-relevant}. This difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:\textbf{we need to reason about (the computation about) extensible pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}. When proving "subst_lemma", we have to know "subst" is invariant on "tt", i.e. \mintinline{Coq}{(subst i x tt) = tt}. This kind of information requires \textit{computational rules} from the recursors.
% we want tactic programming and certain level of automation
Finally, \textlabel{Req (5)}{langdesign-req5}:\textbf{we desire tactic programming} to relieve us from the hassle of direct manipulation of proof terms.
